### 2017-07-05 ThesisText/background.tex

E-Learning is usually provided over the Internet using platforms
called~\gls{lms}. Examples of~\gls{lms} include Moodle, Blackboard Learn, D2L,
among others. These platforms organize users by separating them into roles such
as Professors and Students or Authors and Learners. The Professors or Authors
are the users who organize courses in universities or training in companies,
create and publish content, perform evaluation of students, etc. Students or
Learners are the users which actually perform the training and follow courses
in the platforms.

@@

The contents of a course may be resources such as PDF files, links, media such
as images, videos, etc. These resources should be guides for students achieve
their objectives in learning, or should be resources needed for the student to
perform certain tasks. A course may also have activities. Activities may be,
for example, projects that a student needs to complete before a given deadline.
Activities may also be quizzes made to students over Moodle. These quizzes may
be used for evaluation.

@@

One difficulty of clustering is to find the number of clusters that optimally
divide the data. Some algorithms take a number of clusters as an arguments,
while others are able to find an optimal value.

By it's own nature, it may not be possible to find clusters in data, that is,
there may not be any great division of objects given their features in a given
dataset. After executing clustering algorithms a performance measure is used to
determine if that division is good or not.

@@

K-Means is an clustering algorithm that find $ K $ clusters in data. The $ K $
variable is given to the algorithm as an argument, hence the name K-Means. The
algorithm works by creating $ K $ random points based on the features of the
data. The algorithm then iterates updating the location of those points until
they converge.~\cite{book_ml,book_dm_practical}

@@

The model used in this project and described in this section always go through
two processes. Training and testing. To do this two sets of data are needed,
one for each process. Data for the two processes is usually taken from the same
dataset in different proportions. For example, some projects may take 75\% or
90\% of the data to do training, and use the remainder to test.

@@

The main objective of~\gls{edm} is to extract meaningful information from
educational data like web logs. It is not trivial to do so without advance Data
Mining techniques. \gls{edm}, therefore, aims at training models based on
education data for a variety of tasks.

One such example of an~\gls{edm} tasks is to predict student grades based
features from the students themselves, the courses they are undertaking, and
their activity in a~\gls{lms}. By training such models, professors can make
decisions about how courses are organized regarding students success. For
example, a certain number of activities in a given course may be beneficial to
it's success because it keeps students engaged, but any number of activities
more might not make a difference.~\cite{ind_001, ind_002, ind_005}

@@

In~\cite{ind_001}, some models are trained in order to predict student grades.
The tasks described in the paper are classification tasks, in which a student
is assigned to a given class corresponding to their success in courses based on
their usage of a~\gls{lms} platform. The classes available are Excellent, Good,
Pass, and Fail.

@@

In~\cite{ind_006}, networks are trained from Moodle logs. The data from the
logs are structured in a slightly different way than the previous studies. The
features of this data are shown in table~\ref{tab:ind_006_fields}.

>>>>

E-Learning is usually provided over the Internet using platforms
called~\gls{lms}. Examples of~\gls{lms} include Moodle, Blackboard Learn, D2L,
among others. These platforms organize users by separating them into roles such
as Professors and Students or Authors and Learners. The Professors or Authors
are the users who organize courses in universities or training in companies,
create and publish content, perform evaluation of students, etc. Students or
Learners are the users who actually perform the training and follow courses in
the platforms.

@@

The contents of a course may be resources such as PDF files, links, media such
as images, videos, etc. These resources should be guides for students to
achieve their objectives in learning, or should be resources needed for the
student to perform certain tasks. A course may also have activities. Activities
may be, for example, projects that a student needs to complete before a given
deadline.  Activities may also be quizzes made to students over Moodle. These
quizzes may be used for evaluation.

@@

One difficulty of clustering is to find the number of clusters that optimally
divide the data. Some algorithms take a number of clusters as an argument,
while others are able to find an optimal value.

By its own nature, it may not be possible to find clusters in data, that is,
there may not be any great division of objects given their features in a given
dataset. After executing clustering algorithms a performance measure is used to
determine if that division is good or not.

@@

K-Means is a clustering algorithm that find $ K $ clusters in data. The $ K $
variable is given to the algorithm as an argument, hence the name K-Means. The
algorithm works by creating $ K $ random points based on the features of the
data. The algorithm then iterates updating the location of those points until
they converge.~\cite{book_ml,book_dm_practical}

@@

The model used in this project and described in this section always go through
two processes, training and testing. To do this, two sets of data are needed,
one for each process. Data for the two processes is usually taken from the same
dataset in different proportions. For example, some projects may take 75\% or
90\% of the data to do training, and use the remainder to test.

@@

The main objective of~\gls{edm} is to extract meaningful information from
educational data like web logs. It is not trivial to do so without advanced
Data Mining techniques. \gls{edm}, therefore, aims at training models based on
education data for a variety of tasks.

One such example of an~\gls{edm} task is to predict student's grades based
features from the students themselves, the courses they are undertaking, and
their activity in a~\gls{lms}. By training such models, professors can make
decisions about how courses are organized regarding students success. For
example, a certain number of activities in a given course may be beneficial to
its success because it keeps students engaged, but any number of activities
more might not make a difference.~\cite{ind_001, ind_002, ind_005}

@@

In~\cite{ind_001}, some models are trained in order to predict student's
grades.  The tasks described in the paper are classification tasks, in which a
student is assigned to a given class corresponding to their success in courses
based on their usage of a~\gls{lms} platform. The classes available are
Excellent, Good, Pass, and Fail.

@@

In~\cite{ind_006}, networks are trained from Moodle logs. The data from the
logs is structured in a slightly different way than the previous studies. The
features of this data are shown in table~\ref{tab:ind_006_fields}.

### 2017-07-05 ThesisText/data.tex

The rest of the sheets contain the objects which are going to be used in the
Data Mining tasks. Those objects and their features are fairly unorganized in
the original data, so an analysis on of the available features had to be done.
Section~\ref{sec:objs_feat} states the objects found in the data and details
their features. Section~\ref{sec:preproc_dataset} shows how the final dataset
is organized.

@@

The logs contain the actions of the Moodle users which role can be of a
professors or a students. The data contains features for Students but no
features for professors. Table~\ref{tab:students_features}, contains all the
features of students.

@@

From table~\ref{tab:students_features} it is observable that a student may be
identified by it's internal Moodle Student ID, or by it's student number as
given by SIIUE. Some students in this data repository do not have a
\textit{StudentsNumber} value, but those students also don't have any activity
in the logs.

@@

The fields \textit{Results} and \textit{FinalResult} are always empty together
and in that cases there is never any grade.

@@

Each Moodle user has a username, an actual name, and a role. That information
is stored in \texttt{MoodleUsers}. This dataset contains users which are both
professors and students. They are identified by their role.

>>>>

The rest of the sheets contains the objects which are going to be used in the
Data Mining tasks. Those objects and their features are fairly unorganized in
the original data, so an analysis on of the available features had to be done.
Section~\ref{sec:objs_feat} states the objects found in the data and details
their features. Section~\ref{sec:preproc_dataset} shows how the final dataset
is organized.

@@

The logs contain the actions of the Moodle users which role can be of a
professor or a student. The data contains features for students but no
features for professors. Table~\ref{tab:students_features}, contains all the
features of students.

@@

From table~\ref{tab:students_features} it is observable that a student may be
identified by its internal Moodle Student ID, or by its student number as
given by SIIUE. Some students in this data repository do not have a
\textit{StudentsNumber} value, but those students also don't have any activity
in the logs.

@@

The fields \textit{Results} and \textit{FinalResult} are always empty together
and in those cases there is never any grade.

@@

Each Moodle user has a username, an actual name, and a role. That information
is stored in \texttt{MoodleUsers}. This dataset contains users who are both
professors and students. They are identified by their role.

### 2017-07-05 ThesisText/experiments.tex

\subsection{Predicting Students Grades Based on Moodle Usage}
\label{sec:exp_001}
% EXP_001

A first experiment on this dataset aims at predicting the students grades based
on their Moodle usage data and in their other characteristics present on this
dataset. In this experiment six tasks are made. Both Decision Trees and Naive
Bayes Classifiers are trained using the dataset with binary classes, classes
with intervals of two, and classes with intervals of four.

@@

    \caption
        [Predicting students grades based on Moodle usage results]
        {Predicting students grades based on Moodle usage results. Two models
        are trained 10 times on three different variations of the same dataset.
        The maximum, minimum, and average accuracy is displayed for those ten
        executions.}

@@

\subsection{Clustering Students Grades Based on Moodle Usage}
\label{sec:exp_002}
% EXP_002

@@

    \caption
        [Clustering Students Grades Based on Moodle Usage]
        {Clustering Students Grades Based on Moodle Usage. The table displays
        the results for the silhouette score for clustering. The values for
        every variant of the dataset are approximately the same, so only one
        table is shown.}

@@

Table~\ref{tab:exp_002a_res} has a listing of objects that are placed in each
of the two clusters. It is observable that the major difference is in the
number of~\gls{crud} activities. Different values for activities remain the
same for different courses and enrollment numbers, but greater values are
placed in one cluster while lower values are placed in another.

@@

Some classification tasks use the \texttt{ApprovedCourses} feature as a class.
In this case, feature may be left in it's original form or it may be replaced
by classes. There are three ways to do this replacement.

@@

The results shown were obtain with the K-Means algorithm. Affinity Propagation
was also executed, but the results obtained were not good, having scores within
0.0 and 0.5.

@@

In this dataset we relate the profiling of a student in a courses and his
results on that course. To begin constructing this dataset we use the Profiling
dataset, as shown in section~\ref{sec:profile}.

@@

\subsection{Predicting Students Grades From Student/Course Profiling}
\label{sec:exp_005}
% EXP_005

Using the described dataset, we train models to predict student grades based on
profiling data. Table~\ref{tab:exp_005_res} shows the results of six tasks.
Three for each Model. Each tasks is done in a variant of the dataset.

@@

    \caption
        [Predicting students grades from student/course profiling results]
        {Predicting students grades from student/course profiling results. The
        results show the maximum, minimum, and average for each trained model
        and each dataset variant.}

@@

\subsection{Clustering Students Grades From Student/Course Profiling}
\label{sec:exp_006}
% EXP_006

@@

    \caption
        [Clustering Students Grades From Student/Course Profiling]
        {Clustering Students Grades From Student/Course Profiling. The values
        for every variant of the dataset are approximately the same, so only
        one table is shown.}

>>>>

\subsection{Predicting Student's Grades Based on Moodle Usage}
\label{sec:exp_001}
% EXP_001

A first experiment on this dataset aims at predicting the student's grades based
on their Moodle usage data and in their other characteristics present on this
dataset. In this experiment six tasks are made. Both Decision Trees and Naive
Bayes Classifiers are trained using the dataset with binary classes, classes
with intervals of two, and classes with intervals of four.

@@

    \caption
        [Predicting student's grades based on Moodle usage results]
        {Predicting student's grades based on Moodle usage results. Two models
        are trained 10 times on three different variations of the same dataset.
        The maximum, minimum, and average accuracy is displayed for those ten
        executions.}

@@

\subsection{Clustering Student's Grades Based on Moodle Usage}
\label{sec:exp_002}
% EXP_002

@@

    \caption
        [Clustering Student's Grades Based on Moodle Usage]
        {Clustering Student's Grades Based on Moodle Usage. The table displays
        the results for the silhouette score for clustering. The values for
        every variant of the dataset are approximately the same, so only one
        table is shown.}

@@

Table in attachment~\ref{attach:exp_002} has a listing of objects that are
placed in each of the two clusters. It is observable that the major difference
is in the number of~\gls{crud} activities. Different values for activities
remain the same for different courses and enrollment numbers, but greater
values are placed in one cluster while lower values are placed in another.

@@

Some classification tasks use the \texttt{ApprovedCourses} feature as a class.
In this case, feature may be left in its original form or it may be replaced by
classes. There are three ways to do this replacement.

@@

The results shown were obtained with the K-Means algorithm. Affinity
Propagation was also executed, but the results obtained were not good, having
scores within 0.0 and 0.5.

@@

In this dataset we relate the profiling of a student in a course and his
results on that course. To begin constructing this dataset we use the Profiling
dataset, as shown in section~\ref{sec:profile}.

@@

\subsection{Predicting Student's Grades From Student/Course Profiling}
\label{sec:exp_005}
% EXP_005

Using the described dataset, we train models to predict student's grades based
on profiling data. Table~\ref{tab:exp_005_res} shows the results of six tasks,
three for each Model. Each task is done in a variant of the dataset.

@@

    \caption
        [Predicting student's grades from student/course profiling results]
        {Predicting student's grades from student/course profiling results. The
        results show the maximum, minimum, and average for each trained model
        and each dataset variant.}

@@

\subsection{Clustering Student's Grades From Student/Course Profiling}
\label{sec:exp_006}
% EXP_006

@@

    \caption
        [Clustering Student's Grades From Student/Course Profiling]
        {Clustering Student's Grades From Student/Course Profiling. The values
        for every variant of the dataset are approximately the same, so only
        one table is shown.}

### 2017-07-05 ThesisText/final.tex
Corrected typos

In this project it was shown how the original data from the University of
Évora's Moodle was preprocessed into a data repository useful for analysis.
Some statistics over that data were then shown. The experiments were broken
down by dataset. Each dataset was build from the pre processed original data.

@@

\begin{enumerate}
    \item Predict student grades from their usage of Moodle;
        (Section~\ref{sec:exp_001})
    \item Predict how many courses a student is going to complete given his
        usage of Moodle and initial number of enrolled courses;
        (Section~\ref{sec:exp_003})
    \item Predict student grades from his usage of features in Moodle.
        (Section~\ref{sec:exp_005})
\end{enumerate}

@@

The results of an~\gls{edm} study may have many applications. Experiments such
as the ones made to dataset~\ref{sec:exp_001_002} and~\ref{sec:exp_005_006}
yield insight into how the basic usage of Moodle is a great indicator of the
success of students. From the trained models, future applications may be
develop which look into students and their usage of Moodle and finally predict
their success even before a student as finished a course. With such
applications we can detect student behaviour that might lead to his failure of
a course and signal that student to take action to avoid failure.

From the Moodle data described in this project, further classification and
clustering analysis may also be done. Mode Datasets may be built and analyse from
the original data and more algorithms may be used. For example, a dataset may
be built that joins Moodle usage data (as described in
sections~\ref{sec:exp_001_002} and~\ref{sec:exp_005_006}) and profiling data
(as described in section~\ref{sec:exp_003_004}). A dataset with all these
fields might yield better accuracies in the prediction of grades or more ways
to do clustering.

>>>>

In this project it was shown how the original data from the University of
Évora's Moodle was preprocessed into a data repository useful for analysis.
Some statistics over that data were then shown. The experiments were broken
down by dataset. Each dataset was build from the preprocessed original data.

@@

\begin{enumerate}
    \item Predict student's grades from their usage of Moodle;
        (Section~\ref{sec:exp_001})
    \item Predict how many courses a student is going to complete given his
        usage of Moodle and initial number of enrolled courses;
        (Section~\ref{sec:exp_003})
    \item Predict student's grades from his usage of features in Moodle.
        (Section~\ref{sec:exp_005})
\end{enumerate}

@@

The results of an~\gls{edm} study may have many applications. Experiments such
as the ones made to dataset~\ref{sec:exp_001_002} and~\ref{sec:exp_005_006}
yield insight into how the basic usage of Moodle is a great indicator of the
success of students. From the trained models, future applications may be
developed which look into students and their usage of Moodle and finally
predict their success even before a student as finished a course. With such
applications we can detect student behaviour that might lead to his failure of
a course and signal that student to take action to avoid failure.

From the Moodle data described in this project, further classification and
clustering analysis may also be done. More datasets may be built and analysed
from the original data and more algorithms may be used. For example, a dataset
may be built that joins Moodle usage data (as described in
sections~\ref{sec:exp_001_002} and~\ref{sec:exp_005_006}) and profiling data
(as described in section~\ref{sec:exp_003_004}). A dataset with all these
fields might yield better accuracies in the prediction of grades or more ways
to do clustering.
